Supplementary Report for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2022-10-03 experiments/2_supplement/supplement.Rmd
changes to supplement

We made new simulations with communities composed of functional groups all containing only the strain with the greatest tolerance. E.g., in a simulation to be compared with variation in the only the SB group (and no variation in the CB or PB groups) then the CB and PB groups (as before) each had only one strain with the mean tolerance value, whereas the SB group contained one strain with higher tolerance. (We also performed simulations in which we retained 9 strains in each functional group, but set the tolerance of all nine to the maximum tolerance value. Results were the same as those presented below.)

For most of the explored combinations and amounts of functional diversity the results are the same (Figure \@ref(fig:max-tol1)). There are differences when there is high functional diversity / high tolerance in the SB, in the CB-SB, and in the CB-SB-PB combinations.  We look at these more closely below.

@@

## SB only, high diversity

With only the most tolerance SB strain, the position of the oxic to anoxic tipping point at high SB diversity appears to be at very low oxygen diffusivity, compared to when all nine strains are present (Figure \@ref(fig:max-tol1)). A corresponding simulations from which the position of the tipping points are calculated are shown in Figures \@ref(fig:max-tol2) (nine strains) and \@ref(fig:max-tol3) (most tolerant strain only). We see that the effect of oxygen diffusivity on the stable state is quite different in the case of only the most tolerant strain only--at low oxygen diffusivity there is variation around a stable state with all three functional groups coexisting--this never occured when there were multiple strains varying in tolerance, or when there was only one strain with mean tolerance. Furthermore, the state of the system is independent of the direction of change in oxygen diffusivity. I.e. there is no region of bistability (no alternate stable states).

We now wish to return to an aspect of the reviewer comment: It is correct that in Fig. 4 when oxygen diffusivity increased, the two dominant functional groups (the sulfur bacteria groups) are dominated by the most tolerant strain before the state transition. And when oxygen diffusivity decreased, the dominant functional group (the cyanobacteria) is dominated by the most tolerant strain before the state transition. It is worth noting, however, that the transitions that occur are to states dominated by the least tolerant-highest growth rate strain. This raises the possibility that less tolerant strains may influence the results.

Indeed, in the SB case, absence of the high growth rate SB strain precludes the formation of a stable state that does not include cyanobacteria. It seems that the most tolerant SB strain does not have a high enough growth rate to itself exclude the cyanobacteria. Cyanobacteria are therefore always present, regardless of the oxygen diffusivity. The stable states of the system then display a complex relationship with oxygen diffusivity that our algorithm for detecting tipping points is not well suited to, hence the rather odd positions of the indicated tipping points.

(ref:max-tol2) Stable states found with nine strains of SB and high diversity. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:max-tol3) Stable states found with the single most tolerant SB and high diversity. All other aspects are the same as in Figure \@ref(fig:repl-meth2). As mentioned in the text, the algorithm used to detect the position of the tipping points is not working correctly here, and so the grey arrows should be ignored.

@@

In the ecosystem shown in Figure \@ref(fig:max-tol3), where there was quite high tolerance SB strain, there was large variation in state variables along the oxygen diffusivity gradient at values less than about -4.5. To examine what might be causing this, we simulated the ecosystem at a constant oxygen diffusivity of $log_{10}$ oxygen diffusivity of -6 (Figure \@ref(fig:SB-maxtol-dyn1)). We see that the dynamics are qualitatively different from other cases in which there is a stable state reached in such a simulation. It appears that the high tolerance SB strain is able to initiate a transition to an anoxic state, but is not able to create a stable anoxic stable state. This dynamical situation was not presented in the ecosystem in which high tolerance and low tolerance strains were present.

@@

(ref:SB-maxtol-dyn1) Temporal dynamics of an ecosystem with a single SB strain with high tolerance, the same one as in Figure \@ref(fig:max-tol3) simulated with constant $log_{10}(oxygen diffusivity) of -6.

@@

## CB-SB only, high diversity

@@

(ref:max-tol5)  Stable states found with the single most tolerant CB strain and SB strain and high diversity. All other aspects are the same as in Figure \@ref(fig:repl-meth2). As mentioned in the text, the algorithm used to detect the position of the tipping points is not working correctly here, and so the grey arrows should be ignored.

>>>>

To test if the diversity effects on resilience were driven entirely by the most tolerant strains, we made additional simulations where functional groups contained only the strain with greatest tolerance and lowest maximum growth rate. That is, for each of the 20 levels of trait variation, we ran simulations where each functional group had only one strain, and its trait values corresponded with those of the most tolerant strain for the given level of trait variation. We also manipulated in which of the functional groups tolerance increased. For example, in a simulation to be compared with variation in only the SB group (and no variation in the CB or PB groups), the SB group contained the strain with highest tolerance, whereas CB and PB each contained the strain with mean tolerance. We then compared the tipping point positions resulting from this tolerance manipulation with the tipping point positions resulting from the diversity manipulation presented in the main text.

To exclude that differences between the tolerance manipulation and the diversity manipulation were driven by the difference in number of strains, we also performed simulations in which we retained nine strains in each functional group but set the tolerance of all nine strains to the maximum tolerance value. Results were the same as those presented below from simulations with one strain.

For most of the explored combinations and amounts of functional diversity/tolerance the tipping point positions did not differ between the diversity manipulation and the tolerance manipulation (Figure \@ref(fig:max-tol1)). However, the tipping point positions did differ between the diversity manipulation and the tolerance manipulation when diversity/tolerance in the SB group was high (i.e. in the combinations SB only, CB-SB, and CB-SB-PB).

@@

## High tolerance vs. high diversity in only the SB

When the SB contained only one strain with very high tolerance and low maximum growth rate, there were no alternate stable states, i.e. the state of the system was independent of the direction of change in oxygen diffusivity (Figure \@ref(fig:max-tol3)). Furthermore, at low oxygen diffusivity the system varied around a stable state with all three functional groups coexisting –this never occurred when there were multiple strains varying in tolerance (Figure \@ref(fig:max-tol2)), or when there was only one strain with mean tolerance.

Apparently, absence of SB strains with high growth rate precludes the formation of a stable state that does not include cyanobacteria. It seems that the most tolerant SB strain does not have a high enough growth rate to itself exclude the cyanobacteria. Cyanobacteria are therefore always present, regardless of the oxygen diffusivity.

(ref:max-tol2) Relationship of ecosystem state variables and oxygen diffusivity when there is variation among nine strains of SB and the amount of variation is high. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:max-tol3) Relationship of ecosystem state variables and oxygen diffusivity when SB contained only the most tolerant strain, with trait values corresponding with the diversity level of the previous figure. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

In the ecosystem shown in Figure \@ref(fig:max-tol3), where the SB contained one strain with quite high tolerance, there was large variation in state variables along the oxygen diffusivity gradient at values less than about -4.5. To examine what might be causing this, we simulated the ecosystem at a constant oxygen diffusivity of $log_{10}$ oxygen diffusivity of -6 (Figure \@ref(fig:SB-maxtol-dyn1)). We see that the dynamics are qualitatively different from other cases in which there is a stable state reached in such a simulation (Figure \@ref(fig:sb-div-effect4)). It appears that the high tolerance SB strain is able to initiate a transition to an anoxic state, but is not able to create a stable anoxic state. This dynamical situation was not presented in the ecosystem in which high tolerance and low tolerance strains were present.

@@

(ref:SB-maxtol-dyn1) Temporal dynamics of an ecosystem with a single SB strain with high tolerance, the same one as in Figure \@ref(fig:max-tol3) simulated with constant $log_{10}$ oxygen diffusivity of -6.

@@

## High diversity vs. high tolerance in CB and SB

@@

(ref:max-tol5)  Stable states found with the single most tolerant CB strain and SB strain and high diversity. All other aspects are the same as in Figure \@ref(fig:repl-meth2).
